---
title: "Using the STAC catalogue to load and aggregate land cover classes"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("gdalcubes")
library("rstac")
library("tibble")
library("sp")
library("sf")
library("dplyr")
library("rgbif")
library("tidyr")
library("stars")
library("stacatalogue")
```


### Loading observations

Let's download [Glyptemys insculpta](https://en.wikipedia.org/wiki/Wood_turtle) observations from GBIF.

```{r}
obs <- rgbif::occ_data(scientificName = "Glyptemys insculpta", hasCoordinate = TRUE, limit = 100)$data
obs <- CoordinateCleaner::cc_val(obs, lon = "decimalLongitude", 
                                 lat = "decimalLatitude", verbose = T, value = "clean")

obs <- CoordinateCleaner::cc_zero(obs, lon = "decimalLongitude", 
                                        lat = "decimalLatitude", buffer = 0.5, 
                                        verbose = T, value = "clean")
obs <- dplyr::select(obs, decimalLongitude, decimalLatitude) %>%
  dplyr::rename(lon = decimalLongitude) %>%
  dplyr::rename(lat = decimalLatitude) 

```

We reproject the decimal longitude and latitude to a user-specified projection system. This conversion to a projected system is essential to interact with the cube.

```{r}
srs.obs <-"EPSG:4326" # initial observations projection system
srs.cube <- "EPSG:6623" # targeted projected projection system
obs.coords.proj <- create_projection(obs, lon = "lon", lat = "lat", 
                                       srs.obs, srs.cube)
head(obs.coords.proj )
```

### Filter classes
                         
Let\'s select crop areas (class 11) and load it at the native resolution (250m) from 2000 to 2001 (2 years). We set prop to FALSE.
```{r, warning=FALSE}
plot_categorical <- function(r, main = "") {
  values <- unique(data.frame(terra::rast(r[[1]]))[,1])
  my_col <- rev(terrain.colors(n = length(values)+1))
  plot(r, legend = FALSE, col = my_col, main = main)
  legend(x = "topright", legend = values, fill = my_col[-1])
}
crop_2000_2001 <-load_prop_values(stac_path = "https://io.biodiversite-quebec.ca/stac/",
                                collections = c("esacci-lc"), 
                                use.obs = T,
                                obs = obs.coords.proj,
                                buffer.box = 0,
                                srs.cube = srs.cube,
                                t0 = "2000-01-01",
                                t1 = "2001-12-31",
                                spatial.res = 250, # in meters
                                prop = F,
                                select_values = c(11),
                                temporal.res =  "P1Y")

```
```{r, warning=FALSE}
plot_categorical(crop_2000_2001$y2000_class11, main = "Crop - year 2000")
```
```{r, warning=FALSE}
plot_categorical(crop_2000_2001$y2001_class11, main = "Crop - year 2001")
```


### Calculate proportion of classes
                         
Let\'s calculate the proportions of crop and urban areas (class 11 and 190) at 1km resolution,  from 2000 to 2001 (2 years). We set prop to TRUE.

```{r, warning=FALSE}
crop_urban_prop <-load_prop_values(stac_path = "https://io.biodiversite-quebec.ca/stac/",
                                collections = c("esacci-lc"), 
                                use.obs = T,
                                obs = obs.coords.proj,
                                buffer.box = 0,
                                srs.cube = srs.cube,
                                t0 = "2000-01-01",
                                t1 = "2001-12-31",
                                spatial.res = 250, # in meters
                                prop = T,
                                prop.res = 1000,
                                select_values = NULL,
                                temporal.res =  "P1Y")

```

```{r, warning=FALSE}
par(mfrow=c(1,2))
plot(crop_urban_prop$y2000_class11, main = "Proportion of crop - 2000")
plot(crop_urban_prop$y2001_class11, main = "Proportion of crop - 2001")
```

```{r, warning=FALSE}
par(mfrow=c(1,2))
plot(crop_urban_prop$y2000_class190, main = "Proportion of urban area - 2000")
plot(crop_urban_prop$y2001_class190, main = "Proportion of urban area - 2001")
```